NMRlipids Databank: Overlay Databank of Lipid Membrane Simulations Arising from Open Collaboration

### 2022-10-05 main.tex
Update text

Here we present a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  The concept of overaly databank is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it pontentially used for wide range of situations, particularly when storage of raw data requires significant resources and final outcomes or best practices are not yet clear, overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.

@@

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Design of the NMRlipids databank}
The practical challenges related to the distribution of large MD simulation trajectories are solved in the NMRlipids databank using the overlay structure. 
The key idea of the overlay databank is that the storage of raw data in layer 1 is distributed in publicly available repositories or other servers with long term stability and permanent links such as digital object identifiers. The core of the databank, layer 2, contains only information on the location and content of the raw data, thereby not requiring large resources to handle and maintain. This lowers the barrier for starting such databank as well as for long term storage. The NMRlipids databank is essentially a git repository containing information on the location of raw data and its indexing with universal naming convention. In addition to all computers where the databank is developed and used, the NMRlipids databank git is stored to Zenodo server, thereby enabling a very cost effective long term storage for the databank. The databank can be used in layer 3 by accessing the raw data and information stored in the databank by employing the universal naming conventions for atoms, molecules and simulation details. The applications can be linked to the core databank (layer 2) without actually including them, thereby enabling flexible extension of the data without compromising the simplicity and lightness of the core databank.

@@

Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

\subsection{Quality evaluation of force fields}
Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19}, but universal quality measure for membrane simulations is not defined and controversial results are often reported from simulations~\cite{antila22b}. The ambiguity in defining the quality of simulations complicates the selection of proper simulation parameters and estimation of the reliability of simulation results, thereby being a major obstacle in many applications of membrane MD simulations. The NMRlipids
databank enables rapid quality evaluation of MD simulations using the quantitative quality measure for membrane simulations. The conformational ensembles of individual lipid molecules are evaluated against C-H bond order parameters from NMR experiments, which is complement by x-ray scattering form factors arising from overall membrane dimensions via electron density profile. Both of these are robust experimental measurables that can be directly connected to the simulation data~\cite{ollila16}.

%NMRlipids databank contains also experimental order parameter and x-ray form factor data that is connected to corresponding  simulations in order to define the simulation qualities and rank them to select the suitable simulation models for particular applications. 
The top quality simulations based on average order parameter quality are shown in Fig.~\ref{fig:quality} A and the direct comparison between the best simulation (POPC with OPLS3e) and experiments in shown in Fig.~\ref{fig:quality} D. This model reproduces the experimental data very well with the exception of third carbon in the $\it{sn}$-1 chain and carbons 8 and 9 close to double bond in $\it{sn}$-2 chain.
%Full details of the NMRlipids quality measure for lipid bilayer simulations are given in the methods section and supplementary information.
The top ranking simulation for the order parameter quality of POPE lipid are shown in Fig.~\ref{fig:quality} B. In this case the best overall quality is found in GROMOS-CKP simulation (direct comparison shown in Fig.~\ref{fig:quality} E, but clear discrepancies with experiments are present in this model for glycerol backbone and $\it{sn}$-2 chain. On the other hand, glycerol backbone and headgroup for POPE is best described in CHARMM36 model (direct comparison shown in Fig.~\ref{fig:quality} F), but its overall quality suffers from overestimated order in acyl chains. Scatter plots and Pearson correlation coefficients in Fig.~\ref{fig:quality} C reveal strong correlation between membrane area per lipid and thickness. The area per lipid is also strongly correlated with the average order of $\it{sn}$-1 chain, suggesting that its order parameters can used as a proxy for the area per lipid in addition to the conformational ensemble. On the other hand, the first form factor minima is correlated stronger with membrane thickness (Fig.~\ref{fig:quality} C) than with the area per lipid (Fig.~\ref{fig:QualityCorrelationsSI}).

@@

The power of NMRlipids quality metrics in selecting the best model for particular application is demontrated in Fig.~\ref{fig:quality} c) where the area per lipids of POPC:POPE mixtures from different force fields are shown and quality of POPE simulation in different force fields are evaluated. Among the simulations available in the NMRlipids databank, the Slipids force field gives the best quality in terms of acyl chain order parameters and form factor, and predicts the largest area per molecule for POPE and smallest difference with POPC. In conclusion, simulations with Slipids is the most reliable force field for membrane packing in POPC:POPE mixtures althought its glycerol backbone is not accurately modelled. Similar comparisons utilizing the preliminary data from the NMRlipids databank have concluded that Slipids is relatively good model also for mixtures of charged POPC:POPS membranes, although models with better counterion binding have higher quality, but CHARMM36 is the best to study differences in headgroup conformational ensembles between different lipid types. Understanding such complex picture of lipid bilayer MD simulation quality would not be possible without automatic quality evaluation of large sets of simulations enabled by the NMRlipids databank.

@@

>>>>

Here we present a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  The concept of overlay databank is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. Particularly when storage of raw data requires significant resources and final outcomes or best practices are not yet clear, overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.

@@

%Up to three levels of \textbf{subheading} are permitted. Subheadings should not be numbered.
\subsection{Overlay structure of the NMRlipids databank}
The three layer structure of the NMRlipids databank, illustrated in Fig.~\ref{fig:overlay} a, enables an efficient way to share and upcycle large MD simulation trajectories.  Layer 1 contains raw simulation data that can locate in any publicly available repository or server offering long term stability and permanent links to the data, such as digital object identifiers. Layer 2 contains all the relevant information on the simulations, including links to the raw data in layer 1 and relevant metadata describing the systems. Layer 3 is the application layer containing outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}) and results from analyses desribed in sections below.

Layer 2 is the core of the databank. In addition to the metadata and links to raw simulation data, it contains universal naming conventions for the molecules and atoms in the databank, computer programs to generate and analyse the databank, basic properties calculated from all simulations (area per lipid, C-H bond order parameters, x-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored is the git repository which is currently available at~\url{https://github.com/NMRLipids/Databank}.
Detailed description on its content is in the supplementary information. Applications in layer 3 uses the information in layer 2 to access raw data in layer 1 to perform automatic analyses over large sets of simulations in the databank.

%are enabled by the 
%The databank can be used in layer 3 by accessing the raw data and information stored in the databank by employing the universal naming conventions for atoms, molecules and simulation details. The applications can be linked to the core databank (layer 2) without actually including them, thereby enabling flexible extension of the data without compromising the simplicity and lightness of the core databank.

@@

\subsection{Content of the databank}
Currently the databank is composed of approximately 500 trajectories with the total length of approximately 231 microseconds of which most are contributed for the previous publications from the NMRlipids open collaboration~\cite{botan15,catte16,antila19,bacle21}. The distribution of lipids, force fields, length of the trjajectories and available binary mixtures are shown in Fig.~\ref{fig:overlay}.

\subsection{Quality evaluation of force fields}
Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19}, but universal quality measure for membrane simulations is not defined and controversial results are often reported from simulations~\cite{antila22b}. The lack of universal quality measure for membrane simulations complicates the selection of proper simulation parameters and the estimation of reliability in simulations, thereby being a major obstacle in many applications of membrane MD simulations. To enable the rapid quality evaluation of membrane MD simulations, a quantitative quality measure based on C-H bond order parameters from NMR experiments is defined in the NMRlipids databank. %Conformational ensembles of individual lipid molecules are evaluated against C-H bond order parameters from NMR experiments. 
This is complemented by comparing the locations of x-ray scattering form factor minima against experiments. 
%that are related to the overall membrane dimensions via electron density profile. 
Both of these are robust experimental measurables that can be directly connected to the simulation data~\cite{ollila16}. The used quality measures are defined in detail in the supplementary information.

%NMRlipids databank contains also experimental order parameter and x-ray form factor data that is connected to corresponding  simulations in order to define the simulation qualities and rank them to select the suitable simulation models for particular applications. 
Simulations with the highest scores based on order parameter evaluation for all simulations are shown in Fig.~\ref{fig:quality} A and for POPE lipids in Fig.~\ref{fig:quality} B. The direct comparison with experiments for the best simulation, POPC with OPLS3e, is shown in Fig.~\ref{fig:quality} C, where discrepancies with experiments are observed only for the third carbon in the $\it{sn}$-1 chain, and carbons 8 and 9 close to the double bond in $\it{sn}$-2 chain. The best overall quality for POPE is found in GROMOS-CKP simulation, for which the direct comparison in Fig.~\ref{fig:quality} D show minor differences in acyl chains, but major discrepancies in glycerol backbone and in the first carbon of $\it{sn}$-2 chain. These parts are better described for POPE by CHARMM36 parameters (direct comparison shown in Fig.~\ref{fig:quality} E), but its overall quality suffers from the overestimated order in acyl chains.

To exemplify the finding of optimal model for a certain lipid, the top ranking simulation for the order parameter quality of POPE lipid are shown in Fig.~\ref{fig:quality} B.

Differences in the predictions of area per lipids for POPC and POPE bilayers between different force fields are demonstrated in Fig.~\ref{fig:POPCPOPEapls}. Because area per lipid is strongly correlated with the average order of $\it{sn}$-1 chain ((Fig.~\ref{fig:quality} F)), its order parameters can used as a proxy for the area per lipid to evaluate its quality against experiments. Simulations with the largest area per lipids for POPE in Fig.~\ref{fig:POPCPOPEapls}, GROMOS-CKP and Slipids, have significantly higher qualities in Fig.~\ref{fig:quality} B than lipid17 and CHARMM36 with smaller are per lipids.  
Among the simulations available in the NMRlipids databank, the Slipids force field gives the best quality in terms of acyl chain order parameters and form factor, and predicts the largest area per molecule for POPE and smallest difference with POPC. In conclusion, simulations with Slipids is the most reliable force field for membrane packing in POPC:POPE mixtures althought its glycerol backbone is not accurately modelled. Similar comparisons utilizing the preliminary data from the NMRlipids databank have concluded that Slipids is relatively good model also for mixtures of charged POPC:POPS membranes, although models with better counterion binding have higher quality, but CHARMM36 is the best to study differences in headgroup conformational ensembles between different lipid types.

On the other hand, the first form factor minima is correlated stronger with membrane thickness (Fig.~\ref{fig:quality} C) than with the area per lipid (Fig.~\ref{fig:QualityCorrelationsSI}). 
Scatter plots and Pearson correlation coefficients in Fig.~\ref{fig:quality} C reveal strong correlation between membrane area per lipid and thickness.

\begin{figure}[tbp]
    \centering
    \includegraphics[width=88mm]{Figures/APL.pdf}
    \caption{Differences in area per lipids of POPC and POPE bilayers between different force fields.  
    }
    \label{fig:POPCPOPEapls}
\end{figure}

Understanding such complex picture of lipid bilayer MD simulation quality would not be possible without automatic quality evaluation of large sets of simulations enabled by the NMRlipids databank.

%The power of NMRlipids quality metrics in selecting the best model for particular application is demontrated in Fig.~\ref{fig:quality} c) where the area per lipids of POPC:POPE mixtures from different force fields are shown and quality of POPE simulation in different force fields are evaluated.

@@

@@

, thereby not requiring large resources to handle and maintain. This lowers the barrier for starting such databank as well as for long term storage. The NMRlipids databank is essentially a git repository containing information on the location of raw data and its indexing with universal naming convention. In addition to all computers where the databank is developed and used, the NMRlipids databank git is stored to Zenodo server, thereby enabling a very cost effective long term storage for the databank.
